---
title: "Syntax-and-Code-in-rMarkdown"
description: |
  rMarkdown Series #1 - r-markdown is an extremely useful tool to generate 
  academic articles. While you could write plain text, you are also able to 
  insert code chunks and generate some results to enrich the file. This 
  article will introduce the mostly used, basic but useful syntax and coding 
  in rMarkdown. 
author:
  - name: Zhuotian Tang
    url: https://theaaaa.github.io/
    affiliation: Marcus Institute for Aging 
    affiliation_url: https://www.marcusinstituteforaging.org/
  - name: Hao Zhu
    url: https://haozhu233.github.io/
    affiliation: Marcus Institute for Aging 
    affiliation_url: https://www.marcusinstituteforaging.org/
  - name: Thomas Travison
    affiliation: Marcus Institute for Aging 
    affiliation_url: https://www.marcusinstituteforaging.org/
date: 11-30-2018
output:
  radix::radix_article:
    self_contained: false
preview: images/hex-rmarkdown.png
draft: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro to R Markdown

R Markdown is a file format for making dynamic documents with R. An R Markdown 
document is written in markdown (a simple formatting syntax for authoring HTML, 
PDF, and MS Word documents) and contains chunks of embedded R code. For more 
details on using R Markdown see <http://rmarkdown.rstudio.com>.

By starting a new R Markdown file, you will firstly be asked to decide the 
title, author's name, and basically the output format of the file. Take the 
**Document** option as an example. HTML format would be a great choice if 
you only need a presenting file, since you could switch to PDF or HTML file 
anytime. 

When you click the **Knit** button on the left top of the window, a document 
will be generated that includes both content as well as the output of any 
embedded R code chunks within the document. Meanwhile, each report contains 
the code it needs to make its own graphs, tables, numbers, etc. The author can 
automatically update the report by re-knitting.



## Markdown Syntax

The text in an R Markdown document is written with the Markdown syntax, or 
precisely speaking, Pandoc's Markdown, which is the most comprehensive one. 

Generally, you could type as usual to write a plain text without any formatting, 
or enter two spaces to start a new paragraph. 

### Text formatting

To make the content *italic*, quote it by a pair of single asterisk marks (* ) 
or underlines(_ ), like `*italic*` or `_italic_`; to make the content **bold**, 
quote it by a pair of double asterisk marks or underlines, like `**bold**` or 
`__bold__`; or maybe you want to make it _**italic and bold**_ at the same time, 
then use any combination of the corresponding symbols mentioned above, e.g., 
`_**italic and bold**_`; 

To mark text as `inline code`, quote it by a pair of backticks, like ``code``. 

A pair of tildes (~ ) turn text to a subscript, while a pair of carets (^ ) 
produce a superscript. For example, `H~3~PO~4~` renders H~3~PO~4~, and `Cu^2+^` 
will render Cu^2+^. 

### Insert links and images

Hyperlinks are created using the syntax `[text](link)`, e.g., 
`[RStudio](https://www.rstudio.com)`. The syntax for images is similar: just 
add an exclamation mark, e.g., `![alt text or image title](path/to/image)`. 
Footnotes are put inside the square brackets after a caret `^[footnotes]`, 
e.g., `^[This is a footnote.]`.

### Section formatting

Section headers can be written after a number of pound signs (# ), while the more 
the pound signs are, the smaller the header size is. The number of pound signs 
could be from one to six, while the size of header with six pound signs is even 
smaller than that of plain text. 

Unordered list items start with *, -, or +, and you can nest one list within 
another list by indenting the sub-list by four spaces, e.g.,

`- one item`  
`- one item`  
`- one item`  
`    - one more item`  
`    - one more item`  
`    - one more item`  

The output will be:

- one item
- one item
- one item
    - one more item
    - one more item
    - one more item

Blockquotes are written after >, e.g.,

`> "I thoroughly disapprove of duels. If a man should challenge me,
  I would take him kindly and forgivingly by the hand and lead him
  to a quiet place and kill him."`  
`>`  
`> --- Mark Twain`  

The output will be:

> "I thoroughly disapprove of duels. If a man should challenge me,
  I would take him kindly and forgivingly by the hand and lead him
  to a quiet place and kill him."
>
> --- Mark Twain

In R code chunk, the pound signs create comments. If you want to make 
comments in document text, use `<!-- regular html comment --> `. In this way, 
then the text quoted would not be shown in rendered document, only veiwable in 
the original rmd file. 

In general, you’d better leave at least one empty line between adjacent but 
different elements, e.g., a header and a paragraph. This is to avoid ambiguity 
to the Markdown renderer. 

### Math expressions

Inline LaTeX equations can be written in a pair of dollar signs using the LaTeX 
syntax, e.g., `$f(k) = {n \choose k} p^{k} (1-p)^{n-k}$` (actual output: 
$f(k) = {n \choose k} p^{k} (1-p)^{n-k}$); 

math expressions of the display style can be written in a pair of double dollar 
signs, e.g., `$$f(k) = {n \choose k} p^{k} (1-p)^{n-k}$$`, and the output looks 
like this: 
$$f(k) = {n \choose k} p^{k} (1-p)^{n-k}$$. 

You can also use math environments inside `$ $` or `$$ $$`, e.g.,

`$$\begin{array}{ccc}`  
`x_{11} & x_{12} & x_{13}\\`  
`x_{21} & x_{22} & x_{23}`  
`\end{array}$$`  

$$\begin{array}{ccc}
x_{11} & x_{12} & x_{13}\\
x_{21} & x_{22} & x_{23}
\end{array}$$  

`$$\begin{vmatrix}a & b\\`  
`c & d`  
`\end{vmatrix}=ad-bc$$`  

$$\begin{vmatrix}a & b\\
c & d
\end{vmatrix}=ad-bc$$  



## R Code Chunks

By clicking on the **Insert** button on the right top of the window, you can 
easily add a code chunk to the file, when there are several kinds of 
programming languages provided. Take an R code chunk as an example. In a code 
chunk, there are lots of things you could do, such as to produce any kinds of 
output, or manipulate a set of data. 

You have fine control over all these output via chunk options, which can be 
provided inside the curly braces (between `{r` and `}`). For example, you can 
choose to hide text output via the chunk option `results = 'hide'`, or set the 
figure height to 4 inches via `fig.height = 4`. Chunk options are separated by 
commas, e.g., `{r, chunk-label, results='hide', fig.height=4}`.

### Chunk options

The value of a chunk option can be an arbitrary R expression, which makes 
chunk options extremely flexible. For example, the chunk option eval controls 
whether to evaluate (execute) a code chunk, and you may conditionally evaluate 
a chunk via a variable defined previously, e.g.,

```{r, echo=TRUE}
# execute code if the date is later than a specified day
do_it = Sys.Date() > '2018-02-14'
```

```{r, echo=TRUE, eval=do_it}
x = rnorm(100)
```

There are a large number of chunk options in knitr documented at 
https://yihui.name/knitr/options. Some most useful ones are listed below:

**eval**: Whether to evaluate a code chunk.

**echo**: Whether to echo the source code in the output document (someone may 
not prefer reading your smart source code but only results).

**results**: When set to `'hide'`, text output will be hidden; when set to 
`'asis'`, text output is written “as-is”, e.g., you can write out raw Markdown 
text from R code (like `cat('**Markdown** is cool.\n')`). By default, text 
output will be wrapped in verbatim elements (typically plain code blocks).

**collapse**: Whether to merge text output and source code into a single code 
block in the output. This is mostly cosmetic: `collapse = TRUE` makes the 
output more compact, since the R source code and its text output are displayed 
in a single output block. The default `collapse = FALS`E means R expressions 
and their text output are separated into different blocks.

**warning**, **message**, and **error**: Whether to show warnings, messages, 
and errors in the output document. Note that if you set `error = FALSE`, 
`rmarkdown::render()` will halt on error in a code chunk, and the error will 
be displayed in the R console. Similarly, when `warning = FALSE` or 
`message = FALSE`, these messages will be shown in the R console.

**include**: Whether to include anything from a code chunk in the output 
document. When `include = FALSE`, this whole code chunk is excluded in the 
output, but note that it will still be evaluated if `eval = TRUE`. When you 
are trying to set `echo = FALSE`, `results = 'hide'`, `warning = FALSE`, 
and `message = FALSE`, chances are you simply mean a single option 
`include = FALSE` instead of suppressing different types of text output 
individually.

### Using packages

Just as like writing code and using packages in an R Script file, you are able 
to use any packages in a code chunk when necessary, and some packages are 
extremely useful here when generating such files. 

For instance, if you want to generate a figure:

```{r}
plot(cars, pch = 18)
```

```{r}
library(plotly)
plot_ly(cars, x = cars$speed, y = cars$dist)
```

```{r}
library(ggplot2)
ggplot(cars, mapping = aes(speed, dist)) +
  geom_point() +
  geom_path()
```

For another example, if you want to create tables in the file, the easiest way 
is to use `knitr::kable()`, which can create tables for HTML, PDF and Word 
outputs, e.g.,

```{r tables-mtcars}
knitr::kable(iris[1:5, ], caption = 'A caption')
```

If you are looking for more advanced control of the styling of tables, you are 
recommended to use the kableExtra package, which provides functions to 
customize the appearance of PDF and HTML tables, e.g.,

```{r}
library(kableExtra)

knitr::kable(iris[1:5, ], caption = 'A caption') %>%
  kable_styling(latex_options = c("striped"), font_size = 16)
```


