---
title: "8 verbs in tidyverse"
description: |
  Tidyverse Series #1 - Data transformation and summarization are the foundation
  of any data analyses. In our R community, fortunately, we have a whole set of
  tools in tidyverse to facilitate this process. This article will be the 
  first of a series of tidyverse tutorials that we are using for our 
  internal training. It covers the 8 fundamental verbs related to data 
  transformation and summarization in dplyr and tidyr. 
author:
  - name: Hao Zhu
    url: https://haozhu233.github.io/
    affiliation: Marcus Institute for Aging 
    affiliation_url: https://www.marcusinstituteforaging.org/
  - name: Thomas Travison
    affiliation: Marcus Institute for Aging 
    affiliation_url: https://www.marcusinstituteforaging.org/
date: 11-13-2018
output:
  radix::radix_article:
    self_contained: false
preview: images/tidyverse-default.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

This post is the first of a series of `tidyverse` tutorials we are using for our staff training. It will be introduce the basic idea of how to use `tidyverse` to do data transformation and analyses (more specifically, it's talking about the major functions in `dplyr` and `tidyr`). In the end, you will learn how to generate a summary stats table using the tools provided in `tidyverse`.

# "tidy data"
Before we get started, I would like to spend a little time on the concept of tidy data. I think even if you are not using R (python, matlab or even Excel are always available there), everyone who needs to deal with data should read this [tidy data](http://vita.had.co.nz/papers/tidy-data.pdf) paper. If you haven't read it before, please stop reading this blog post right now and start to read that paper instead (don't forget to come back :P). Basically, Hadley was saying even for the same piece of information, there are a number of ways to organize the data differently. Some formats might be good for human to process the information but for machines and for data science languages like R or python, the best way to organize data is the "tidy" format, where:

- Each variable has a column
- Each observation has a row

# Introduction to `tidyverse`
The term `tidyverse` was announced in 2016 when Hadley Wickham used it to replace the community-invented term "hadleyverse". It includes a collection of R packages that you will be using everyday. Core packages in `tidyverse` includes `dplyr`, `tidyr`, `ggplot2`, `readr`, `purrr`, etc.. You can refer to their [home page](https://www.tidyverse.org/packages/) for details. Basically, the package `tidyverse` was created for everyone's convenience. In the past, you need to install/load a bunch of packages from this collection. Now you only need to install/load one. 

In this post, we will only talk about `dplyr` and `tidyr` to get you started. 

# 8 verbs in `tidyverse`
`Tidyverse` is easy to learn if you remember the following 8 verbs. 

From `dplyr`:

- `select`: select out **columns**.
- `filter`: filter out **rows**.
- `mutate`: add/edit existing **columns**.
- `summarize`: generate a **new** table/vector that summarize the input.
- `group_by`: perform the above actions (& other actions) **by grouping variables**. 

From `tidyr`:

- `gather`: transform data from **wide to tall** (imagine you grab a bunch of branches).
- `spread`: transform data from **tall to wide** (imagine you release your hand and the branches falls off).
- `separate`: **separate** one column into two/several by symbols(default) or others.

Out of these 8 verbs, there are 3 pairs that does similar kind of things as showing below. `group_by` and `separate` are kind of stand alone. I found it's easy to remember them in this way. 

- select & filter: reduce size.
- mutate & summarize: generate new data or summary.
- gather & spread: data transformation.

Once you mastered these 8 verbs, you could say you can handle at least 80% of your normal tasks. 

# Examples

In this post, we will use `iris` as an example. 

```{r}
library(rmarkdown)
library(kableExtra)
options(kableExtra.html.bsTable = T)  

qtable <- function(x) {
  kable(x) %>%
    kable_styling(c("striped", "hover", "condensed"), full_width = F,
                  position = "left")
}

paged_table(iris, options = list(rows.print = 6))
```

## `select` & `filter`
```{r}
library(tidyverse)

iris %>%
  select(Sepal.Length, Petal.Length, Species) %>%
  filter(Sepal.Length > 6) %>%
  paged_table(options = list(rows.print = 6))
```

I'm sure you have noticed this `%>%` symbol. Technically, it means it will "take the result from left and send it to the function on the right side or below as the first argument". If this explanation confuses you, I found it might be easier to treat it as `r shiny::icon("arrow-right")` or pronounce it as "next" when you read it out. The whole point of using this symbol is to make your code more literal and readable. 


For `select`, there are three helper functions in `dplyr` that worth knowing. They are `starts_with`, `ends_with` and `contains`. We can use them to quickly find out the columns we need.

```{r}
iris %>%
  select(contains("Length"), Species) %>%
  paged_table(options = list(rows.print = 6))
```

Note that you can also use the minus sign `-` in select to indicate that you don't want an item. Also, you can use column index (numbers) instead of column names to do the work. For example, `iris %>% select(-Species)` and `iris %>% select(-5)` both work. 

## `mutate` & `summarize`

It's always good to keep in mind that `mutate` add/edit columns while `summarize` destories the entire table and create new ones. Note that everything in `mutate` is doing vector calculation so for conditional-logics, it's easy to combine it with the power of `ifelse`. (See [here](https://stackoverflow.com/questions/34005423/is-if-faster-than-ifelse) for a comparison between `if` and `ifelse`).

```{r}
iris %>%
  select(Petal.Length) %>%
  mutate(
    Petal.Length2 = Petal.Length + 1,
    Petal.Length3 = ifelse(Petal.Length >= 1.5, TRUE, FALSE)
  ) %>%
  paged_table(options = list(rows.print = 6))
```

```{r}
iris %>%
  summarize(
    Petal.Length.mean = mean(Petal.Length),
    Petal.Length.sd = sd(Petal.Length)
  ) 
```

## `group_by` + `summarize`
`summarize` only shines when you combine it with `group_by`.

```{r}
iris %>%
  group_by(Species) %>%
  summarize(
    Petal.Length.mean = mean(Petal.Length),
    Petal.Length.sd = sd(Petal.Length)
  ) %>%
  qtable()
```

## `group_by` + `filter`
`group_by` also works well with `filter`. For example, this example below with filter out the 2nd row in each category of species. 

```{r}
iris %>%
  group_by(Species) %>%
  filter(row_number() == 2) %>%
  qtable()
```
## `gather` & `spread` 

`gather` and `spread` follows the same syntax: `gather(data, key, value)`, `spread(data, key, value)`. In `gather`, you define `key` and `value` by yourself and your selected data will be organized into these two columns. In `spread`, you tell spread the name of the "key" and "value" you are planning to spread and it will turn the data into the wide format. Note that you need to make sure every row has a unique identifier when you spread. Otherwise you will likely see an error message.

```{r}
iris %>%
  # Note that here I'm removing Species from the gathering list because this 
  # Species column is a factor and I don't want it mess up the column type 
  # of "value".
  # Also here instead of "key", I like to call it "var" because practically 
  # it makes more sense. You can call whatever you want in `gather`. 
  gather(var, value, -Species) %>% 
  paged_table(options = list(rows.print = 6))
```

When you spread the results from `gather`, you will get an Error message because there is no unique identifier.

```{r, eval=FALSE}
iris %>%
  gather(var, value, -Species) %>%
  spread(var, value)
# Error: Duplicate identifiers for rows (301, 302, 303, 304, 305, 
# 306, 307, 308, 309, 310, 311, 312, 313, 314, 315, 316, 317, 318, 
# ...
```

If you add row number as ID, things will be different. 

```{r}
iris %>%
  mutate(id = row_number()) %>%
  gather(var, value, -(Species:id)) %>%
  spread(var, value) %>%
  paged_table(options = list(rows.print = 6))
```

## `gather` + `separate` + `spread` 

It may seem meaningless to gather and then spread but you will see the point if you add `separate` in the pipeline. As you can see in the example below, we split the `var` column we got previously into two columns. By default, `separate` will split columns by any symbols but you can always write customized `regular expression` or position index.

```{r}
iris %>%
  mutate(id = row_number()) %>%
  gather(var, value, -(Species:id)) %>%
  separate(var, into = c("loc", "var")) %>%
  paged_table(options = list(rows.print = 6))
```

After you have another possible "key" column, your spreading part start to become very useful. 

```{r}
iris %>%
  mutate(id = row_number()) %>%
  gather(var, value, -(Species:id)) %>%
  separate(var, into = c("loc", "var")) %>%
  spread(loc, value) %>%
  paged_table(options = list(rows.print = 6))
```

## `gather` + `group_by` + `summarize`

By combining the power of `tidy data` and our `dplyr` verbs, we can generate complex analyses in a few rows of code.

```{r}
iris %>%
  gather(var, value, -Species) %>%
  group_by(Species, var) %>%
  summarise(stat = paste0(
    round(mean(value), 1), " (", round(sd(value), 1), ")"
  )) %>%
  paged_table(options = list(rows.print = 6))
```

After some quick conversion, we can easily turn them into some printable table.

```{r}
iris %>%
  gather(var, value, -Species) %>%
  group_by(Species, var) %>%
  summarise(stat = paste0(
    round(mean(value), 1), " (", round(sd(value), 1), ")"
  )) %>%
  spread(var, stat) %>%
  qtable()
```

# Ending Note
There are many other useful tools in `dplyr` and `tidyr` that were not covered by this intro. For example, the `lag` and `lead` function in `dplyr` can help you easily calculate the difference between rows... Next time, we are going to check out how to combine the tools we learned today with `purrr` and `broom` to quickly generate nice tabulated results from models. 


